Supplementary material to: Empirical Bayes estimation of the normal inverse Gaussian prior in simultaneous-equation models

### 2019-05-13 docs/supplement1.Rnw
changed some nomenclature and added nigig model

% load packages
<<packages, include=FALSE, echo=FALSE>>=
library(cambridge)
@

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE)
@

@@

	\section{Variational Bayes derivations}\label{sec:vbderivations}
	In the following all expectations are with respect to the variational posterior $Q$.
	\subsection{Evidence lower bound}
	We monitor convergence through the evidence lower bound (ELBO). To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
	\begin{align*}
	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
	\end{align*}

	In the inverse Gamma setting, we have:
	\begin{align*}
	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
	\end{align*}
  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models. Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.

@@

	\section{Evidence lower bound}
	We monitor convergence through the evidence lower bound (ELBO). To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
	\begin{align*}
	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
	\end{align*}

	In the inverse Gamma setting, we have:
	\begin{align*}
	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
	\end{align*}
  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models. Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.

  \subsection{Efficient computation}
  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$. In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO. The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$. Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting. Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
  \begin{align*}
  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
  \end{align*}
  where $v_i$ are the singular values of $\X$. The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

  \section{Inverse Gamma model}
  In addition to the inverse Gaussian model, we investigate a generalisation of 
  the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, 
  where we draw the $\gamma_d^2$ from an inverse Gamma distribution:
	$$
	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2),
	$$
	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
	have rescaled the inverse Gamma parameters and reparametrised both 
	distributions as generalized inverse Gaussians:
	\begin{align*}
  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
  \end{align*}
	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.

	\begin{align*}
  q(\bm{\beta}_d) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
  \end{align*}
	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively. Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model. The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
  \end{align*}
  Here, we set
  \begin{align}
  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
  c_d^{(h)} & = \begin{cases*}
  b_d^{(h)} & in the conjugate setting \\
  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
  \end{cases*} \nonumber
  \end{align}
  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.

	<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
digauss <- function(x, theta, lambda, eta) {
  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
}

>>>>

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
knit_hooks$set(document=function(x) {
  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
@

@@

	\section{NIG prior}\label{sec:nigprior}
	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
	and kurtosis:
	\begin{align*}
	  \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d, \\
	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \frac{3\theta_d}{\lambda_d} + 3.
	\end{align*}

	\section{NIGIG prior}\label{sec:nigigprior}
	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
	and kurtosis:
	\begin{align*}
	  \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d^c \theta_j^z  
    \(r_d + \frac{2\theta_d^c}{\lambda_d^c}\), \\
	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}{\lambda_j^z} +
	  3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) + 1}{
	  \lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
	\end{align*}
	with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$.

	\section{Student's \texorpdfstring{$t$}{t} prior}\label{sec:studentstprior}
	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
	and kurtosis:
	\begin{align*}
	  \V(\beta_{jd} | \sigma_d^2) & = \frac{\sigma_d^2 \lambda_d}{\eta_d - 2}, \\
	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \frac{6}{\eta_d - 4} + 3.
	\end{align*}

	\section{Variational Bayes derivations}\label{sec:vbderivations}
	For clarity we have indexed the variational 
	density functions with their respective parameters, which we omitted in the 
	MD. In the following all expectations are with respect to the 
	variational posterior $Q_d$. The variational posterior for $\bbeta_d$ is 
	found as follows:
	\begin{align*}
	  \log q_{\bbeta_d} (\bbeta_d) & \propto \E [\log \mathcal{L}
	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \gamma_d^2, 
	  \sigma_d^2)] \\
	  & \propto - \frac{1}{2} \sum_{i=1}^n 
	  \E \[ \frac{(y_{id} - \x_i \tr \bbeta_d)^2}{\sigma_d^2} \] - 
	  \frac{1}{2} \sum_{j=1}^p 
	  \E \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
	  \mathcal{N}_p (\bmu_d, \bSigma_d), \\
	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
	  [\X \tr \X + \E(\gamma_d^{-2}) \cdot \I ]^{-1}, \\
	  & \text{and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \I ]^{-1} 
	  \X \tr \y_d.
	\end{align*}
	The variational posterior for $\gamma_d^2$ is given by:
	\begin{align*}
	  \log q_{\gamma_d^{2}}(\gamma_d^{2}) & \propto \E 
	  [\log \pi (\bbeta_d | \gamma_d^{2}, \sigma_d^2)] + \log \pi(\gamma_d^{2}) \\
	  & \propto -\frac{p}{2} \log \gamma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - 
	  \frac{3}{2} \log \gamma_d^{2} - 
	  \frac{\lambda_d (\gamma_d^2 - \theta_d)^2}{2 \theta_d^2} \gamma_d^{-2} \\
	  & \propto \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - 
	  \frac{\lambda_d}{2\theta_d^2} \gamma_d^2 - 
	  \frac{1}{2}\[ \lambda_d + \E(\sigma_d^{-2}) \sum_{j=1}^p \E(\beta_{jd}^2)\] 
	  \gamma_d^{-2} , \\
	  q_{\gamma_d^{2}}(\gamma_d^{2}) & \overset{D}{=} \mathcal{GIG} 
	  \(-\frac{p+1}{2}, \frac{\lambda_d}{\theta_d^2}, \delta_{d} \), \\
	  & \text{with } \delta_{d} =\E(\sigma_d^{-2})
	  \left\{\E(\bbeta_{d} \tr) \E(\bbeta_{d}) + \trace[\V(\bbeta_{d})]\right\} + 
	  \lambda_d.
	\end{align*}
	Lastly, we have the variational posterior for $\sigma_d^2$:
	\begin{align*}
	  \log q_{\sigma_d^{2}}(\sigma_d^{2}) & \propto \E 
	  [\log \mathcal{L}(\y_d | \bbeta)d, \sigma_d^2 )] + 
	  \E [\log \pi (\bbeta_d | \gamma_d^2, \sigma_d^2)] + 
	  \log \pi (\sigma_d^{2}) \\
	  & \propto -\frac{n}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{i=1}^n 
	  \frac{\E [(y_{id} - \x_i \tr \bbeta_d)^2]}{\sigma_d^{2}} - 
	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
	  \E \( \frac{\beta_{jd}^2)}{\gamma_d^{-2}}\) \sigma_d^{-2} - 
	  \frac{3}{2} \log \sigma^2_d \\
	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} - 
	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
	  \E(\gamma_d^{-2})\sum_{j=1}^p \E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
	  \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_{d} \), \\
	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr \mathbf{y}_d -
    2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
    \trace [ \X \tr \X \V(\bbeta_d)] + 
    \E(\bbeta_d \tr)  \X \tr \X \E(\bbeta_d ) \\
    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
    \E(\gamma_d^{-2}) \trace [ \V(\bbeta_d)] + 
    \E(\gamma_d^{-2}) \E(\bbeta_d \tr) \E(\bbeta_d ) \bigg\}.
	\end{align*}

	\section{Evidence lower bound}\label{sec:evidencelowerbound}
	The variational Bayes posterior is found through maximisation of an evidence
	lower bound (ELBO). Additionally, we may monitor convergence through this 
	ELBO. To compute the ELBO after updating the EB parameters from 
	$\theta_d^{(l)}, \lambda_d^{(l)}$ to 
	$\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute 
	$\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with:
	\begin{align*}
	  \text{ELBO}_d^{(l + 1)} = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - 
	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + 
	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + 
	  \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
	  & - \frac{p + n + 1}{2} \log \zeta_d^{(l)} + 
	  \frac{p + 1}{4} \log \lambda_d^{(l)} - 
	  \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
	  & + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} 
	  \delta_d^{(l)}}/\theta_d^{(l)}\) + 
	  \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
	  & + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + 
	  \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} 
	  e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - 
	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
	\end{align*}
	where $a^{(l)}$, $b^{(l)}$, and $e^{(l)}$ are as in the MD. To compute the 
	ELBO after a VB update we simply let 
  $\theta_d^{(l)}, \lambda_d^{(l)}=\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.

@@

  \section{Efficient computation}
  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$. In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO. The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$. Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting. Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
  \begin{align*}
  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
  \end{align*}
  where $v_i$ are the singular values of $\X$. The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

  \section{Ratios of modified Bessel functions}\label{sec:ratiosmodifiedbessels}
  Ratios of modified Bessel functions of the second kind 
  $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large
  $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
  large, this causes numerical issues in the calculation of various quantities.
  We alleviate the numerical issues through the following.

  \section{Gibbs sampler}\label{sec:gibbssampler}
  MCMC samples from the posterior corresponding to model 
  (\ref{md-eq:priormodel}) may be generated for each equation independently. 
  MCMC samples may be generated by iteratively sampling the following 
  conditional distributions:
  \begin{subequations}
    \begin{align}
      \bm{\beta}_d | \gamma_d^2, \sigma_d^2, \y_d & \sim \mathcal{N}_p 
      (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
      \text{with } & \bm{\Sigma}_d = \sigma_d^2 (\X \tr \X + 
      \gamma_d^{-2} \I)^{-1} \label{eq:sigmasample} \\
      \text{and } & \bm{\mu}_d = (\X \tr \X + 
      \gamma_d^{-2} \I)^{-1} \X \tr \y_d \label{eq:musample} \\
      \gamma_d^2 | \bm{\beta}_d, \sigma_d^2, \y & \sim \mathcal{GIG}
      \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \nonumber \\
      \text{with } & \delta_d = \sigma_d^{-2} \bm{\beta}_d \tr \bm{\beta}_d + 
      \lambda_d, \nonumber \\
      \sigma_d^2 | \bm{\beta}_d, \gamma_d^2, \y & \sim \Gamma^{-1} 
      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
      \text{with } & \zeta = \frac{1}{2} \[\mathbf{y} \tr \mathbf{y} -
      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
      (\bm{\mu}) \tr \X \tr \X \bm{\mu}  + 
      \gamma^{-2} \trace ( \bm{\Sigma}) + 
      \gamma^{-2} (\bm{\mu}) \tr \bm{\mu}\]. \nonumber
    \end{align}
  \end{subequations}
  In high dimensional space, the $p \times p$ matrix inversions in 
  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
  a method to sample from (\ref{eq:betasample}) without explicit calculation
  of this inverse, thereby offering a siginifcant speed-up compared to naive
  sampling.

  \section{Inverse Gamma model}\label{sec:inversegamma}
  The model introduced in MD Section \ref{md-sec:inversegammamodel} draws the 
  $\gamma_d^2$ from an inverse Gamma distribution:
	\begin{equation}\label{eq:invgammaprior}
	  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2)=
	  \mathcal{GIG}(-\eta_d/2, 0, \lambda_d),
	\end{equation}
	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
	have rescaled the inverse Gamma parameters and reparametrised as a generalized
	inverse Gaussian. This allows for an easier comparison to the inverse 
	Gaussian hyperprior:
	\begin{equation}\label{eq:invgaussianprior}
	  \mathcal{IG}(\theta_d, \lambda_d) =
	  \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d).
  \end{equation}
	From (\ref{eq:invgammaprior}) and (\ref{eq:invgaussianprior}) it is obvious 
	that the two models coincide for $\theta_d \to \infty$ and $\eta_d = 1$. In
	Figure \ref{fig:dens_igaussian_igamma} the two distributions are compared for 
	a range of parameter values.
<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
digauss <- function(x, theta, lambda, eta) {
  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
}
